Analyzing microtomography data with Python and the scikit-image library

### 2016-10-29 skimage_tomo.tex
Modified section titles to please the publisher
Replace gist citation with Zenodo GH repo
Minor changes following comments from reviewers.

\title{Analyzing X-ray images in Python with scikit-image}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={emmanuelle.gouillart@saint-gobain.com}   % email address
]{\fnm{Emmanuelle} \snm{Gouillart}}
\author[
   addressref={aff2},
   %email={}
]{\fnm{Juan} \snm{Nunez-Iglesias}}
\author[
   addressref={aff3},
   email={stefanv@berkeley.edu}
]{\fnm{Stéfan} \snm{van der Walt}}

@@

Several software applications and libraries are available to synchrotron
users to process their images. ImageJ \citep{Abramoff2004, Schneider2012}
and its distribution Fiji \citep{Schindelin2012} is a popular tool for
2-D and 3-D images, thanks to its intuitive menus and graphical tools,
and the wealth of plugins contributed by a vivid community
\citep{Schindelin2015}. For 3-D images, commercial tools such as Avizo 3D
software (TM) are appreciated for an intuitive graphical pipeline and
advanced 3D visualization. Alternatively, the use of a programming
language gives finer control, better reproducibility, and more complex analysis
possibilities, provided classical processing algorithms can be called
from libraries -- thereby limiting the complexity of the programming task
and the risk of bugs. Matlab (TM) and its image processing toolbox are
popular in the academic community of computer vision and image
processing. The Python language is widely used in the scientific world
and in synchrotron facilities. As a general-purpose language, Python is
used in synchrotrons to control device servers \citep{pytango}, to access
raw data of X-ray detectors \citep{Knudsen2013}, to reconstruct
tomography volumes from radiographs \citep{Gursoy2014, Mirone2014}, and
in data processing packages for macromolecular cristallography
\citep{Adams2010}, azimuthal integration of diffraction data
\citep{Ashiotis2015}, or fluorescence analysis \citep{pymca}.

\texttt{scikit-image} \citep{Vanderwalt2014} is a general-purpose image
processing library for the Python language, and a component of the ecosystem of
Python scientific modules commonly known as Scientific
Python \citep{Oliphant2007}. Like the rest of the ecosystem,
\texttt{scikit-image} is released under a permissive open-source license and is
available free of charge. Most of \texttt{scikit-image} is compatible with both 2-D and
3-D images, so that it can be used for a large number of imaging modalities,
such as microscopy, radiography or tomography. In this article, we explain how
\texttt{scikit-image} can be used for processing data acquired in X-ray
imaging experiments, with a focus on microtomography 3-D images. This
article does not intend to be a pedagogical tutorial on \texttt{scikit-image}
for X-ray imaging, but rather to explain the rationale behind the package, and
provide various examples of its capabilities.

\section*{Overview and first steps}

@@

\begin{figure*}
    \centerline{\includegraphics[width=0.77\textwidth]{ecosystem_landscape}}
    \caption{\csentence{\texttt{scikit-image} and the Scientific Python
	ecosystem}. Images are opened from files as \texttt{NumPy}
	arrays. Functions of \texttt{scikit-image} transform image arrays
	into other arrays with the same dimensions, or into arrays of
	numbers corresponding to features of the image. The output of
	scikit-image functions can be passed to other Python modules
	relying on \texttt{NumPy} arrays, such as \texttt{SciPy} or
	\texttt{scikit-learn}. Image-shaped arrays are transformed into
	visualizations with \texttt{matplotlib} (2D) or \texttt{Mayavi}
	(3D). A variety of environments is available for code development
	and execution, from classical IDEs to Jupyter notebooks.
    \label{fig:ecosystem}}
\end{figure*}

@@

\begin{figure*}[ht!]
    \centerline{\includegraphics[width=0.99\textwidth]{spyder_process}}
    \caption{\csentence{The Spyder IDE} integrates a text editor (with
	syntax highlighting), the IPython interpreter, as well as a panel
	for code introspection (online help, variable explorer, \dots).
 \label{fig:spyder}}
\end{figure*}

\begin{figure*}[ht!]
    \centerline{\includegraphics[width=0.75\textwidth]{jupyter_example}}
    \caption{\csentence{The Jupyter notebook} allows mixing of computer code (top),
             plot and text output (middle), and free-form narrative text (bottom).
             This makes it ideal to record and report code-based analyses.
             Screenshot from \url{https://gist.github.com/jni/3339985a016572f178d3c2f18e27ec0d}.
 \label{fig:jupyter}}
\end{figure*}

@@

\begin{figure}[ht!]
    \centerline{\includegraphics[width=0.6\columnwidth]{mayavi_example}}
    \caption{Simple visualization realized with Mayavi.
 \label{fig:mayavi}}
\end{figure}

@@

@@

\texttt{scikit-image} offers most classical image processing operations,
such as exposure and color adjustment, filtering,
segmentation, feature extraction, geometric transformations, and
measurements of region characteristics. In addition, some advanced
algorithms are also implemented, a selection of which is illustrated in
Fig.~\ref{fig:tomo_gallery}.

Tomographic images often suffer from artifacts or poor signal-to-noise
ratio. Several denoising filters are available for restoring these
images, ranging from general-purpose median and bilateral
filters to those more suited to specific applications. For example, total-variation denoising \citep{Chambolle2004,
Getreuer2012} is ideal for restoring piecewise-constant images (see
Fig.~\ref{fig:tomo_gallery} a)), such as images with a small number of
phases encountered in materials science \citep{Bouttes2015}. Conversely,
images with a fine-grained texture are better preserved with non-local
means denoising \citep{Buades2005} (see Fig.~\ref{fig:tomo_gallery} a)).

\texttt{scikit-image} offers a wide variety of functions for detecting
geometrical features of interest in an image. Peaks in 2D
diffraction images can be extracted using blob detection methods
\citep{Ashiotis2015}, such as the Laplacian of Gaussian method (see
Fig.~\ref{fig:tomo_gallery} b)). In order to detect thin boundaries, the
ridges of an image can be identified as regions for which the leading eigenvalue
of the local Hessian matrix is high (see Fig.~\ref{fig:tomo_gallery} b)).

Segmentation of regions of interest can be achieved using one of various
strategies, depending on the characteristics of the image. Images with a
clear contrast between regions can be segmented automatically thanks to
several thresholding algorithms, including an adaptive local thresholding
algorithm aimed at images with contrast variations. Super-pixel
algorithms \citep{Felzenszwalb2004, Achanta2012} create an
over-segmentation of images in super-pixels, by grouping pixels that are
close together both in color- and spatial distance (see
Fig.~\ref{fig:tomo_gallery} c)). Region-growing algorithms, such as the
morphological watershed or the random walker \citep{Grady2006}, propagate
the labels of user-defined markers through the image (see
Fig.~\ref{fig:tomo_gallery} c)). The active contour algorithm
\citep{Kass1988} fits snake contours to features of the image, such as
edges or high-brightness regions.

The characteristics of labeled regions (particles, porosities, organs,
\dots) resulting from a segmentation can be measured using the
\texttt{measure} submodule. The different connected components (e.g.
bubbles or non-touching particles) of a binary image are labeled with the
\texttt{measure.label} function. Properties of labeled regions such as
size, extent, center of mass or mean intensity value are accessed with
\texttt{measure.regionprops} (see Fig.~\ref{fig:tomo_gallery} d)). Local
characteristics of a region can be retrieved as well:
Fig.~\ref{fig:tomo_gallery} d) shows how the local diameter of open
porosity is measured by combining a skeletonization of the porosity
channels, and the distance transform to the other phase measured on the
skeleton.

@@

\section*{Current limitations and challenges}

@@

\paragraph{Speed of execution.} Although \texttt{scikit-image} approaches
the speed of execution of compiled (C++, Java) code, it cannot reach the
performance of code optimized for the GPU, or the hand-tuned
CPU-specific optimizations found in OpenCV. At the moment,
\texttt{scikit-image} is not the best tool for ultrafast computations
where the workflow is known beforehand, simple and stable. However, it is
an excellent tool for exploring image data interactively and testing
different algorithms -- an important component of data processing in
scientific work --, and its speed of execution is sufficient for processing
gigabyte-sized tomographic images in seconds to minutes. Moreover,
the multiprocessing capability of \texttt{scikit-image} is likely to improve in
the near future.

@@

@@

>>>>

\title{Analyzing microtomography data with Python and the scikit-image
library}

@@

\author[
   addressref={aff1},                   % id's of addresses, e.g. {aff1,aff2}
   corref={aff1},                       % id of corresponding address, if any
   %noteref={n1},                        % id's of article notes, if any
   email={emmanuelle.gouillart@nsup.org}   % email address
]{\fnm{Emmanuelle} \snm{Gouillart}}
\author[
   addressref={aff2},
   %email={}
]{\fnm{Juan} \snm{Nunez-Iglesias}}
\author[
   addressref={aff3},
   email={stefanv@berkeley.edu}
]{\fnm{Stéfan} \snm{van der Walt}}

@@

Several software applications and libraries are available to synchrotron
users to process their images. ImageJ \citep{Abramoff2004, Schneider2012}
and its distribution Fiji \citep{Schindelin2012} is a popular
general-purpose tool for 2-D and 3-D images, thanks to its intuitive
menus and graphical tools, and the wealth of plugins contributed by a
vivid community \citep{Schindelin2015}. Software specialized in analyzing
synchrotron data are available as well, such as XRDUA~\citep{xrdua} for
diffraction images obtained in powder diffraction analysis, or  For 3-D
images, commercial tools such as Avizo 3D software (TM), or
ToolIP/MaviKit~\citep{mavi} are appreciated
for an intuitive graphical pipeline and advanced 3D visualization. Some
synchrotrons have even developed their own tools for volume processing,
such as Pore3D~\citep{Brun2010} at the Elettra facility. Alternatively,
the use of a programming language gives finer control, better
reproducibility, and more complex analysis possibilities, provided
classical processing algorithms can be called from libraries -- thereby
limiting the complexity of the programming task and the risk of bugs.
Matlab (TM) and its image processing toolbox are popular in the academic
community of computer vision and image processing. The Python language is
widely used in the scientific world and in synchrotron facilities. As a
general-purpose language, Python is used in synchrotrons to control
device servers \citep{pytango, Sugandhi2016, nsls-ii}, to access raw data
of X-ray detectors \citep{Knudsen2013}, to reconstruct tomography volumes
from radiographs \citep{Gursoy2014, Mirone2014}, and in data processing
packages for macromolecular cristallography \citep{Adams2010}, azimuthal
integration of diffraction data \citep{Ashiotis2015}, or fluorescence
analysis \citep{pymca, Sole2007}.

%\vskip 1cm 
\texttt{scikit-image}
\citep{Vanderwalt2014} is a general-purpose image processing library for
the Python language, and a component of the ecosystem of Python
scientific modules commonly known as Scientific Python
\citep{Oliphant2007}. Like the rest of the ecosystem,
\texttt{scikit-image} is released under a permissive open-source license
and is available free of charge. Most of \texttt{scikit-image} is
compatible with both 2-D and 3-D images, so that it can be used for a
large number of imaging modalities, such as microscopy, radiography or
tomography. In this article, we explain how \texttt{scikit-image} can be
used for processing data acquired in X-ray imaging experiments, with a
focus on microtomography 3-D images. This article does not intend to be a
pedagogical tutorial on \texttt{scikit-image} for X-ray imaging, but
rather to explain the rationale behind the package, and provide various
examples of its capabilities.

\section*{Methods -- Overview and first steps}

@@

\begin{figure*}[!htb]
    \centerline{\includegraphics[width=0.77\textwidth]{ecosystem_landscape}}
    \caption{\csentence{\texttt{scikit-image} and the Scientific Python
	ecosystem}. Images are opened from files as \texttt{NumPy}
	arrays. Functions of \texttt{scikit-image} transform image arrays
	into other arrays with the same dimensions, or into arrays of
	numbers corresponding to features of the image. The output of
	scikit-image functions can be passed to other Python modules
	relying on \texttt{NumPy} arrays, such as \texttt{SciPy} or
	\texttt{scikit-learn}. Image-shaped arrays are transformed into
	visualizations with \texttt{matplotlib} (2D) or \texttt{Mayavi}
	(3D). A variety of environments is available for code development
	and execution, from classical IDEs to Jupyter notebooks.
    \label{fig:ecosystem}}
\end{figure*}

@@

\begin{figure*}[!htb]
    \centerline{\includegraphics[width=0.99\textwidth]{spyder_process}}
    \caption{\csentence{The Spyder IDE} integrates a text editor (with
	syntax highlighting), the IPython interpreter, as well as a panel
	for code introspection (online help, variable explorer, \dots).
 \label{fig:spyder}}
\end{figure*}

\begin{figure*}[!htb]
    \centerline{\includegraphics[width=0.75\textwidth]{jupyter_example}}
    \caption{\csentence{The Jupyter notebook} allows mixing of computer code (top),
             plot and text output (middle), and free-form narrative text (bottom).
             This makes it ideal to record and report code-based analyses.
             Screenshot from \url{https://gist.github.com/jni/3339985a016572f178d3c2f18e27ec0d}.
 \label{fig:jupyter}}
\end{figure*}

@@

@@

\section*{Results}
\begin{figure}[!htb]
    \centerline{\includegraphics[width=0.6\columnwidth]{mayavi_example}}
    \caption{\csentence{Simple 3-D visualization} realized with Mayavi.
 \label{fig:mayavi}}
\end{figure}

@@

\texttt{scikit-image} offers most classical image processing operations,
such as exposure and color adjustment, filtering, segmentation, feature
extraction, geometric transformations, and measurements of region
characteristics. In addition to common operations, some advanced
algorithms are also implemented, a selection of which is illustrated in
Fig.~\ref{fig:tomo_gallery}. In the following, we briefly illustrate how
\texttt{scikit-image} can be used for some typical image processing
tasks encountered when analyzing tomographic images: denoising, mid-range
feature detection, segmentation and measurement of region properties. For
the sake of brevity, other tasks such as contrast manipulation or
geometric transformations are not described here; the interested reader
is referred to the documentation of \texttt{scikit-image}.

Tomographic images often suffer from artifacts or poor signal-to-noise
ratio. Therefore, \emph{denoising} data is often the first step of an
image processing workflow. Several denoising filters are available for
restoring these images, ranging from general-purpose median and
bilateral filters to those more suited to specific applications. For
example, total-variation denoising \citep{Chambolle2004, Getreuer2012} is
ideal for restoring piecewise-constant images (see
Fig.~\ref{fig:tomo_gallery} a)), such as images with a small number of
phases encountered in materials science \citep{Bouttes2015}. Conversely,
images with a fine-grained texture are better preserved with non-local
means denoising, a patch-based algorithm \citep{Buades2005} (see
Fig.~\ref{fig:tomo_gallery} a)).

Detecting the presence of objects or extracting pixels corresponding to
objects (a task known as segmentation) is an important task of image
analysis for medical or materials science applications.
\texttt{scikit-image} offers a wide variety of functions for
\emph{detecting geometrical features} of interest in an image. In order
to detect thin boundaries, the ridges of an image can be identified as
regions for which the leading eigenvalue of the local Hessian matrix is
high (see Fig.~\ref{fig:tomo_gallery} b)). In the Fourier space, peaks in
2D Bragg diffraction patterns can be extracted using blob detection
methods \citep{Ashiotis2015}, such as the Laplacian of Gaussian method
(see Fig.~\ref{fig:tomo_gallery} b)).

\emph{Segmentation} of regions of interest can be achieved using one of various
strategies, depending on the characteristics of the image. Images with a
clear contrast between regions can be segmented automatically thanks to
several thresholding algorithms, including an adaptive local thresholding
algorithm aimed at images with contrast variations. Super-pixel
algorithms \citep{Felzenszwalb2004, Achanta2012} create an
over-segmentation of images in super-pixels, by grouping pixels that are
close together both in color- and spatial distance (see
Fig.~\ref{fig:tomo_gallery} c)). Region-growing algorithms, such as the
morphological watershed or the random walker \citep{Grady2006}, propagate
the labels of user-defined markers through the image (see
Fig.~\ref{fig:tomo_gallery} c)). The active contour algorithm
\citep{Kass1988} fits snake contours to features of the image, such as
edges or high-brightness regions.

Following segmentation, the characteristics of labeled regions
(particles, porosities, organs, \dots) resulting from a segmentation can
be measured using the \texttt{measure} submodule. The different connected
components (e.g. bubbles or non-touching particles) of a binary image are
labeled with the \texttt{measure.label} function. Properties of labeled
regions such as size, extent, center of mass or mean intensity value are
accessed with \texttt{measure.regionprops} (see
Fig.~\ref{fig:tomo_gallery} d)). Local characteristics of a region can be
retrieved as well: Fig.~\ref{fig:tomo_gallery} d) shows how the local
diameter of open porosity is measured by combining a skeletonization of
the porosity channels, and the distance transform to the other phase
measured on the skeleton.

@@

\section*{Discussion -- Current limitations and challenges}

@@

\paragraph{Speed of execution.} Although \texttt{scikit-image} approaches
the speed of execution of compiled (C++, Java) code, it cannot reach the
performance of code optimized for the GPU, or the hand-tuned
CPU-specific optimizations found in OpenCV \citep{Pulli2012}. At the moment,
\texttt{scikit-image} is not the best tool for ultrafast computations
where the workflow is known beforehand, simple and stable. However, it is
an excellent tool for exploring image data interactively and testing
different algorithms -- an important component of data processing in
scientific work --, and its speed of execution is sufficient for processing
gigabyte-sized tomographic images in seconds to minutes. Moreover,
the multiprocessing capability of \texttt{scikit-image} is likely to improve in
the near future.

@@

\section*{Author's contributions}
E. Gouillart, J. Nunez Iglesias and S. van der Walt conducted the
research and wrote the paper. All authors read and approved the final
manuscript.

@@

\newpage
~
\newpage

\section*{Figures}

\listoffigures
